Where Do Events Come From?

### 2018-08-09 paper.tex
...

\usepackage{listings}
\lstdefinelanguage{ceu}{%
  language=C,
  morekeywords={%
    @const, @pure, @safe, NEVER, FOREVER, PAR, PROC, SIGNAL, abort, and,
    await, bool, call, class, data, define, deterministic, do, each, else,
    emit, end, escape, event, every, finalize, hor, implementation, in,
    input, interface, loop, min, native, new, nohold, not, or, output, par,
    pool, pure, return, s, h, ms, signal, spawn, tag, then, this, traverse, until,
    code, public, private,
    var, watching, when, with},
}
\lstset{
  literate={~} {$\sim$}{1},
  basicstyle=\footnotesize\ttfamily,
  captionpos=b,
  columns=flexible,
  commentstyle=\footnotesize\itshape,
  escapeinside={||},
  frame=tb,
  keepspaces=true,
  keywordstyle=\bfseries,
  language=ceu,
  numbersep=4pt,
  numberstyle=\tiny,
  %xleftmargin=0.5cm,
  %xrightmargin=0.5cm,
  %upquote=true,
}

@@

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment that generates inputs to the application and consumes outputs from
it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
targeting embedded systems in order to take control of the whole event loop,
from inputs to outputs.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to avoid race conditions at
compile time, support lexically-scoped drivers, and provide automatic standby
for applications.
%and hot swapping of device drivers
\end{abstract}

@@

The event-based interface between the reactive language and the environment is
arguably inevitable,
% and also happens at the appropriate layer, since it
%connects the application with the operating system (OS) resources through a
%non-invasive API.
%
%However, this separation
but reveals the environment as a rigid system component
that evolves in separate from the application.
%
It also requires two languages and the programmer has to deal with multiple
syntaxes, incompatible type systems, and different address spaces.
%
Furthermore, in the context of embedded systems, a proper host OS may even be
absent,
% or lacking enough device drivers,
which requires more low-level
intervention from the application.

In this work, we propose interrupt service routines (ISRs) as an asynchronous
primitive for the synchronous reactive language \CEU~\cite{ceu.sensys13}.
ISRs empower reactive applications with their own device drivers which
self-generate inputs, bypassing the need for a host environment.
%
\CEU targets OS-less embedded architectures, such as Arduino-compatible
microcontrollers. %, in which applications run in the bare metal, without
%OS support.
%In this context, device drivers are typically libraries compiled together with
%the main program.
%
%The lack of an OS opens opportunities to explore a tighter integration between
%the application and its device drivers at the language level, resulting in an
%overall simplicity and tractability of the system.
%
%In particular, \CEU is already amenable to a simple static analysis to detect
%race conditions which we extended in this work to also encompass ISRs.
Applications in \CEU can share data buffers with drivers, to avoid unnecessary copying,
with some static guarantee that no data races will occur.
%
\CEU also provides a lexical finalization mechanism % that we adopt in drivers
to
properly disable interrupts and turn off peripherals completely.
%For instance, we can delimit the scope of drivers similarly to lexically-scoped
%variables.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in idle states susceptible to standby.
With the help of drivers and a power manager, the microcontroller
sleeps automatically at optimal sleeping modes after each input reaction.
%
%We implemented an extensible runtime that exposes hooks for the ISRs and
%power manager, which we validated in two microcontrollers: an
%\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, and nRF24L01 transceiver. %, SPI
%
%The applications built on top of these drivers show optimal energy savings due
%to automatic standby.

Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
sensor networks, which also provides asynchronous events with %compile-time
data race detection~\cite{wsn.nesc}, and automatic energy management
~\cite{wsn.icem}.
%
We adapted these ideas to the structured reactive model of \CEU,
and refined them with stronger guarantees due to its support for lexically
scoped lines of execution with automatic finalization.

@@

Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating}. %,rp.rescala,sync_async.cooperative

@@

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost.
%
%The event loop in Figure~\ref{fig.env} suggests this behavior since
The
environment can only generate a new input after the application yields control
and closes the event loop.
%
In Listing~\ref{lst.ceu}, even if two packets arrive simultaneously in the
environment, the synchronous model ensures atomicity and responsiveness in the
application:
    the first \code{await} awakes and turns the LED off atomically
    (ln~12--13);
    and the second \code{await} will awake in sequence (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this might lead to race conditions at a fine grain, but might also
affect the ordering of events at a coarse grain: the effect of an earlier event
might be perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
%Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
%original application represented as the synchronous side and the environment as
%the asynchronous side, all inside the same \CEU application.
%
Our idea is to push all subtleties of asynchronous execution into device
drivers, which emit input events to regular synchronous code in \CEU, exactly
as before.

@@

As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
GPIOs to connect a button (pin 02) to an LED (pin 13) via software such that
the LED is on whenever the button is pressed and off whenever it is unpressed.

Listing~\ref{lst.gpio.async}, the asynchronous side, implements the driver for
the output and input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) with interpolation to
evaluate \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries of embedded
toolchains such as Arduino.
%
In the example, when the driver is included, it configures pin 13 as output
(ln~3) and sets its new state whenever \code{PIN\_13} is emitted (ln~5).
%
An \code{input} event implementation uses an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically perform simple operations and emit an input event
inside its body to awake the synchronous side (ln~17).
However, although the ISR executes asynchronously when the interrupt occurs,
the \code{emit} goes into a queue (to be discussed) and does not interrupt the
ongoing (suspended) reaction on the synchronous side.
%
In the example, the input driver also configures pin 2 to behave as input and
to generate external interrupts on level transitions (ln~11--15).

@@

  output int ADC_REQUEST;
  input  int ADC_DONE;

@@

The next example in Listing~\ref{lst.adc} illustrates the use of a
lexically-scoped driver for the A/D converter.
%
The application is a typical periodic sensor sampling loop, but which
designates an explicit \code{do-end} block (ln~20--24) to include the driver
(ln~21) and use it (ln~22--23).
%
The driver specifies a finalization code (ln~6--12) which executes
unconditionally whenever its enclosing block goes out of scope (ln~24).
The finalize completely disables all A/D functionality to save energy
(ln~8--10).
%
As discussed in the previous section, the finalizer would also execute if
aborted from a hypothetical \code{par/or} enclosing the driver.
%
Note that the driver interface (ln~3--4) is visible only inside the block
(ln~20--24) and thus cannot be inadvertently used outside it.

\subsection{Safe Shared-Memory Concurrency}
\label{sec.ceu.shared}

@@

In \CEU, when multiple trails awake in the same reaction, they execute in
lexical order, i.e., in the order they appear in the source code.
%
In Figure~\ref{lst.mem}, both examples define a shared variable (ln~3) and
assign to it in parallel trails (ln~6, 9).
%
In Listing~\ref{lst.mem.a}, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B} (ln~5,8), which cannot
occur simultaneously in accordance with the synchronous model.
%Hence, for the sequence of events \code{A->B}, \code{x} becomes
%\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
%\code{3}~(\code{(1*2)+1}).
In Listing~\ref{lst.mem.b}, the two assignments to \code{y} are simultaneous
because they execute in reaction to the same event \code{A}.
Nevertheless, because \CEU follows lexical order, the execution is still
deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
%
Even so, \CEU performs (optional) concurrency checks at compile time to detect
conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot access that variable~\cite{ceu.sensys13}.
%
%The static checks are optional (or raise warnings) and do not affect the
%semantics of the language.

@@

\CEU treats a \code{spawn~async/isr} as a block of code that runs in parallel with
the rest of the program (hence the prefix \code{spawn}).
%
This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
Listing~\ref{lst.usart} may lead to a race condition (ln~9,21), and \CEU raises
a compile-time error.
%
The programmer is responsible to protect concurrent memory accesses with
\code{atomic} blocks, which disables interrupts for a short period
of time (supposedly).

@@

@@

However, we do not expect that application programmers should be required to
resolve race conditions.
\CEU supports reactive abstractions (similar to coroutines) that help hiding
the concurrency complexity inside the driver.
%
Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
abstraction that expects a reference to a buffer and a number of bytes to
receive.
Now, the application (ln~29--35) invokes the abstraction (ln~33) passing a
local buffer (ln~32).
%
The driver (ln~5--27) now hides the low level interface of the previous example
(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
access to the shared buffer with an \code{atomic} block (ln~18--21).
%
The abstraction copies the driver buffer into the application buffer (ln~19)
up to the requested number of bytes (ln~22).
%
On the one hand the extra copying incurs a memory and runtime overhead, but on
the other hand it prevents race conditions with the ISR.

@@

\subsection{Energy-Efficient Runtime}
\label{sec.energy}

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Overall runtime architecture of \CEU with an input queue (ln~1),
           event loop (ln~4--11), synchronous execution (ln~7), and standby
           mode (ln~9).},
  label={lst.env},
]
evt_t queue[MAX];           // input queue
void main () {
  ceu_start();              // "boot reaction"
  while (1) {
    evt_t evt;
    if (ceu_input(&evt)) {  // query input queue
      ceu_sync(&evt);       // execute synchronous
    } else {
      ceu_pm_sleep();       // nothing to execute
    }
  }
}
\end{lstlisting}

The language runtime
%implements the event loop proposed in Figure~\ref{fig.async}, which
alternates between synchronous, well-behaved execution, and asynchronous,
unpredictable ISRs.
%
Since the language now has full control over the event loop and only executes
from interrupt requests, it is possible for the runtime to enter in sleep mode
automatically to save energy.
%
Listing~\ref{lst.env} is a realization of this architecture:
%
The runtime defines an input queue (ln~1) in which ISRs enqueue new events
(e.g., Listing~\ref{lst.usart.code}, ln~12).
%
In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
which starts the program, spawns the ISRs, and reaches the first \code{await}
statements in the multiple lines of code of the program.
%
Then, the runtime enters the infinite event loop (ln~4--11) that queries the input
queue (ln~6) to awake the program (ln~7).
%
Note that ISRs execute asynchronously with the loop and
emit an input event into the queue which will be queried in a subsequent
iteration.
%
Note also that if the queue is empty, the runtime enters in sleep mode to save
energy (ln~9), and will only awake on a new hardware interrupt.

@@

The third configuration achieves optimal efficiency but requires a tight
interaction between the drivers and the power management runtime.
%
Listing~\ref{lst.pm} unveils the power manager (ln~1--18), the modified USART
driver (ln~20--29), and an illustrative application (ln~31--43).
%
The application is a loop that initially awaits in parallel for a button press (ln~36) or
receiving 10 bytes (ln~39).
Let's call this state \code{STATE-A}.
The \code{par/or} terminates when either of them occurs, going to the next line
that awaits another button click (ln~42).
Let's call this other state \code{STATE-B}.
After another button click, the program loops back to \code{STATE-A}.
%
While in \code{STATE-A}, the program can awake from USART and external
interrupts, which means that the power manager should choose a sleep mode in
which the USART remains operational.
%
While in \code{STATE-B}, only external interrupts should awake the program,
which means that the power manager may use the most efficient sleep mode.
%
The power manager enumerates all microcontroller's peripherals (ln~3--8) and,
before sleeping, queries their states (ln~11,13) to choose the most appropriate
sleep mode (ln~12,14,16).
%
The USART driver needs to be extended with calls to enable and disable the
USART state inside the power manager: just before awaiting (ln~28), the driver enables
the USART (ln~24) and creates a finalization block to unregister it on
termination (ln~25--27).
Termination may occur either directly after receiving the requested number of
bytes, or indirectly if the \code{par/or} in the application terminates on a
button click (ln~36).

@@

We extend the synchronous language \CEU with asynchronous interrupt service
routines (ISRs) to take control of the whole event loop in reactive systems,
from inputs to outputs.
%
ISRs empower reactive applications to also implement their own device drivers
and self-generate inputs, bypassing the need for a host environment.
%
However, asynchronous execution confronts the well-behaved semantics of
synchronous languages with possible race conditions.
%
To mitigate this threat, we extend the static concurrency checks of \CEU to also
detect data races between ISRs and the application.

>>>>

\usepackage{listings}
\lstdefinelanguage{ceu}{%
  language=C,
  morekeywords={%
    NEVER, FOREVER, and, high, low, async, isr, include, byte, none, atomic,
    await, bool, call, data, define, deterministic, do, else,
    emit, end, escape, event, every, finalize, in,
    input, loop, native, new, nohold, not, or, output, par,
    pool, return, s, h, ms, signal, spawn, tag, then, this, traverse, until,
    code, var, watching, with},
}
\lstset{
  literate={~} {$\sim$}{1},
  basicstyle=\footnotesize\ttfamily,
  captionpos=b,
  columns=flexible,
  commentstyle=\footnotesize\itshape,
  escapeinside={||},
  frame=tb,
  keepspaces=true,
  keywordstyle=\bfseries,
  language=ceu,
  numbersep=4pt,
  numberstyle=\tiny,
  %xleftmargin=0.5cm,
  %xrightmargin=0.5cm,
  %upquote=true,
}

@@

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment that generates inputs to the application and consumes outputs from
it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
targeting embedded systems in order to take control of the whole event loop,
from input generation up to output effects.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to avoid race conditions at
compile time, support lexically-scoped drivers, and provide automatic standby
for applications.
%and hot swapping of device drivers
\end{abstract}

@@

The event-based interface between the reactive language and the environment is
arguably inevitable,
% and also happens at the appropriate layer, since it
%connects the application with the operating system (OS) resources through a
%non-invasive API.
%
%However, this separation
but reveals the environment as a rigid system component
that evolves in separate from the application.
%
It also requires the programmer to deal with multiple
syntaxes, incompatible type systems, and different address spaces.
%
Furthermore, in the context of embedded systems, a proper host operating system (OS) may even be
absent or lacking enough device drivers,
which requires more low-level
intervention from the application.

In this work, we propose interrupt service routines (ISRs) as an asynchronous
primitive for the synchronous reactive language \CEU~\cite{ceu.sensys13}.
ISRs empower reactive applications with their own device drivers which
self-generate inputs, bypassing the need for a host environment.
%
\CEU targets OS-less embedded architectures, such as Arduino-compatible
microcontrollers. %, in which applications run in the bare metal, without
%OS support.
%In this context, device drivers are typically libraries compiled together with
%the main program.
%
%The lack of an OS opens opportunities to explore a tighter integration between
%the application and its device drivers at the language level, resulting in an
%overall simplicity and tractability of the system.
%
%In particular, \CEU is already amenable to a simple static analysis to detect
%race conditions which we extended in this work to also encompass ISRs.
Applications in \CEU can share data buffers with drivers, to avoid unnecessary copying,
with some static guarantee that no data races will occur.
%
\CEU also provides a lexical finalization mechanism % that we adopt in drivers
which
safely disables drivers and turns off peripherals completely to save energy.
%For instance, we can delimit the scope of drivers similarly to lexically-scoped
%variables.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in idle states susceptible to standby.
With the help of drivers and a power manager, the microcontroller
sleeps automatically at optimal sleeping modes after each input reaction.
%
%We implemented an extensible runtime that exposes hooks for the ISRs and
%power manager, which we validated in two microcontrollers: an
%\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, SPI, and an RF transceiver.
%
%The applications built on top of these drivers show optimal energy savings due
%to automatic standby.

Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
sensor networks, which also provides asynchronous events with %compile-time
data race detection~\cite{wsn.nesc} and automatic energy
management~\cite{wsn.icem}.
%
We adapted these ideas to the structured reactive model of \CEU,
and refined them with stronger guarantees due to its support for lexically
scoped lines of execution with automatic finalization.

@@

Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.

@@

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost, which we refer to as the
\emph{atomicity} and \emph{responsiveness} properties, respectively.
%
%The event loop in Figure~\ref{fig.env} suggests this behavior since the
%environment can only generate a new input after the application yields control
%and closes the event loop.
%
In Listing~\ref{lst.ceu}, even if two packets arrive simultaneously in the
environment, the synchronous model ensures these properties by adopting an
event queue:
    the first \code{await} awakes and turns the LED off atomically
    (ln~12--13);
    only after that, the second \code{await} will awake (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this might lead to race conditions at a fine grain, but might also
affect the ordering of events at a coarse grain: the effect of an earlier event
might be perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
%Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
%original application represented as the synchronous side and the environment as
%the asynchronous side, all inside the same \CEU application.
%
Our approach is to push all subtleties of asynchronous execution into device
drivers, which impersonate the environment and are responsible to emit input
events towards regular synchronous code.
%
Synchronous code preserves the atomicity and responsiveness properties,
communicating with the environment only through events, exactly as before.

As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
GPIOs to connect a button to an LED via software such that the LED (pin 13) is
on whenever the button (pin 02) is pressed, and off whenever it is unpressed.

@@

Listing~\ref{lst.gpio.async}, the asynchronous side, implements the driver for
the output and input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) with interpolation to
evaluate \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries of embedded
toolchains such as Arduino.
%
In the example, when the driver is included, it configures pin 13 as output
(ln~3) and sets its new state whenever \code{PIN\_13} is emitted (ln~5).
%
An \code{input} event implementation uses an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically perform simple operations and emit an input event
to awake the synchronous side (ln~17).
However, although the ISR executes asynchronously when the interrupt occurs,
the \code{emit} goes into a queue and does not affect the synchronous side
immediately.
%
In the example, when the input driver is included, it also configures pin 2 to
behave as input and to generate external interrupts on level transitions
(ln~11--15).

@@

  output int ADC_REQUEST;   // requests a conversion
  input  int ADC_DONE;      // conversion is done

@@

The next example in Listing~\ref{lst.adc} illustrates the use of a
lexically-scoped driver for the A/D converter.
%
The application is a typical periodic sensor sampling loop, but which
designates an explicit \code{do-end} block (ln~20--24) to include the driver
(ln~21) and use it (ln~22--23).
%
The driver specifies a finalization code (ln~6--12) which executes
unconditionally whenever its enclosing block goes out of scope (ln~24).
The \code{finalize} completely disables all A/D functionality to save energy
(ln~8--10).
%
As discussed in the previous section, the finalizer would also execute if
aborted from a hypothetical \code{par/or} enclosing the driver.
%
Note that the driver interface (ln~3--4) is visible only inside the block
(ln~20--24) and thus cannot be inadvertently used outside it.

\newpage
\subsection{Safe Shared-Memory Concurrency}
\label{sec.ceu.shared}

@@

In \CEU, when multiple trails awake in the same reaction, they execute
atomically in lexical order, i.e., in the order they appear in the source code.
%
In Figure~\ref{lst.mem}, both listings define a shared variable (ln~3) and
assign to it in parallel trails (ln~6, 9).
%
In Listing~\ref{lst.mem.a}, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B} (ln~5,8), which cannot
occur simultaneously due to the atomicity property of the synchronous model.
%Hence, for the sequence of events \code{A->B}, \code{x} becomes
%\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
%\code{3}~(\code{(1*2)+1}).
In Listing~\ref{lst.mem.b}, the two assignments to \code{y} are simultaneous
because they execute in reaction to the same event \code{A}.
Nevertheless, because \CEU follows lexical order and executes atomically, the
outcome is still deterministic, and \code{y} always becomes \code{4}
(\code{(1+1)*2}).
%
Even so, \CEU performs (optional) concurrency checks at compile time to detect
conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot access that variable~\cite{ceu.sensys13}.
%
%The static checks are optional (or raise warnings) and do not affect the
%semantics of the language.

@@

\CEU treats a \code{spawn~async/isr} as a block of code that runs in parallel with
the rest of the program (hence the prefix \code{spawn}).
%
This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
Listing~\ref{lst.usart} may lead to race conditions (e.g., ln~9,21), and \CEU
raises a compile-time error.
%
The programmer is responsible to protect concurrent memory accesses with
\code{atomic} blocks, which disables interrupts for a short period
of time.

@@

input none USART_RX;    // data is available
var[32] byte rx_buf;    // data buffer

@@

However, we do not expect that application programmers should be required to
resolve race conditions.
\CEU supports reactive abstractions (similar to coroutines) that help hiding
concurrency issues inside drivers.
%
Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
abstraction that expects a reference to a buffer and a number of bytes to
receive.
Now, the application (ln~29--35) invokes the abstraction (ln~33) passing a
local buffer (ln~32).
%
The driver (ln~5--27) now hides the low level interface of the previous example
(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
accesses to the shared buffer with an \code{atomic} block (ln~18--21).
%
The abstraction copies the driver buffer into the application buffer (ln~19)
up to the requested number of bytes (ln~22).
%
On the one hand the extra copying incurs a memory and runtime overhead, but on
the other hand it prevents race conditions with the ISR.

@@

\newpage
\subsection{Energy-Efficient Runtime}
\label{sec.energy}

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Overall runtime architecture of \CEU with an input queue (ln~1),
           event loop (ln~4--11), synchronous execution (ln~7), and standby
           mode (ln~9).},
  label={lst.env},
]
evt_t queue[MAX];           // input queue
void main () {
  ceu_start();              // "boot reaction"
  while (1) {
    evt_t evt;
    if (ceu_input(&evt)) {  // queries input queue
      ceu_sync(&evt);       // executes synchronous
    } else {
      ceu_pm_sleep();       // nothing to execute
    }
  }
}
\end{lstlisting}

The language runtime
%implements the event loop proposed in Figure~\ref{fig.async}, which
alternates between synchronous, well-behaved execution, and asynchronous,
unpredictable ISRs.
%
Since the language now has full control over the event loop and only executes
from interrupt requests, it is possible for the runtime to enter in sleep mode
automatically to save energy.
%
Listing~\ref{lst.env} is a realization of this architecture:
%
The runtime defines an input queue (ln~1) in which ISRs enqueue new events
(e.g., Listing~\ref{lst.usart.code}, ln~12).
%
In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
which starts the program, spawns the ISRs, and reaches the first \code{await}
statements in the multiple trails of the program.
%
Then, the runtime enters the infinite event loop (ln~4--11) that queries the input
queue (ln~6) to awake the program (ln~7).
%
ISRs execute asynchronously, emitting input events into the queue, which will
be queried in subsequent iterations of the loop.
%
Note that if the queue is empty, the runtime enters in sleep mode to save
energy (ln~9), and will only awake on a new hardware interrupt.

@@

The third configuration achieves optimal efficiency but requires a tight
interaction between the drivers and the power manager.
%
Listing~\ref{lst.pm} unveils the power manager (ln~1--18), the modified USART
driver (ln~20--29), and an illustrative application (ln~31--43).
%
The application is a loop that initially awaits in parallel for a button press (ln~36) or
receiving 10 bytes (ln~39).
Let's call this state \code{STATE-A}.
The \code{par/or} terminates when either of them occurs, going to the next line
that awaits another button click (ln~42).
Let's call this other state \code{STATE-B}.
After another button click, the program loops back to \code{STATE-A}.
%
While in \code{STATE-A}, the program can awake from USART and external
interrupts, which means that the power manager should choose a sleep mode in
which the USART remains operational.
%
While in \code{STATE-B}, only external interrupts should awake the program,
which means that the power manager may use the most efficient sleep mode.
%
The power manager enumerates all microcontroller's peripherals (ln~3--8) and,
before sleeping, queries their states (ln~11,13) to choose the most appropriate
sleep mode (ln~12,14,16).
%
The USART driver now needs to be extended with calls to enable and disable the
USART state inside the power manager: just before awaiting (ln~28), the driver enables
the USART (ln~24) and creates a finalization block to disable it on
termination (ln~25--27).
Termination may occur either directly after receiving the requested number of
bytes, or indirectly if the \code{par/or} in the application terminates on a
button click (ln~36).

@@

We extend the synchronous language \CEU with asynchronous interrupt service
routines (ISRs) to take control of the whole event loop in reactive systems,
from input generation up to output effects.
%
ISRs empower reactive applications to also implement their own device drivers
and self-generate inputs, bypassing the need for a host environment.
%
However, asynchronous execution confronts the well-behaved semantics of
synchronous languages with possible race conditions.
%
To mitigate this threat, we extend the static concurrency checks of \CEU to also
detect data races between ISRs and the application.
